---
title: "Microbiome composition for Amplicon data"
author: "Leo Lahti, Sudarshan Shetty et al. `r Sys.Date()`"
bibliography: 
- bibliography.bib
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
    toc_float: true
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    use_bookdown: false
    highlight: haddock    
---
  <!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - composition}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->
  
  
  
Also see [phyloseq barplot examples](http://joey711.github.io/phyloseq/plot_bar-examples.html).
  
Check the [core microbiome page](http://microbiome.github.io/microbiome/CoremicrobiotaAmplicon.html) which shows how to read the your files into R and make a phyloseq object.  

```{r composition-amplicon-example1, warning=FALSE, message=FALSE, eval=TRUE}
# Example data
library(microbiome)
# Try another theme
# from https://github.com/hrbrmstr/hrbrthemes
# you can install these if you don't have it already.
# devtools::install_github("hrbrmstr/hrbrthemes")
library(hrbrthemes)
library(gcookbook)
library(tidyverse)
library(dplyr)
library(jeevanuDB)

ps1 <- emp_human
colnames(tax_table(ps1))
```

In case you see the taxonomic classification is just lablled as "Rank1" ... "Rank7" we can change it as follows  

```{r, composition-amplicon-example1a, eval=TRUE}
# First change the column names of the taxonomy table in phyloseq to following:

colnames(tax_table(ps1)) <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species" )

tax_table(ps1)[tax_table(ps1)[,"Domain"]== "NA", "Domain" ] <- "Unidentified_Domain"

tax_table(ps1)[tax_table(ps1)[,"Phylum"]== "p__", "Phylum" ] <- "p__Unidentified_Phylum"

```

# Composition barplots

The compositon plots can be shown either as barplots or heatmaps. Both examples are show below.  

## Plot counts abundance  

Now we can improve the plot further.  
Let's try at Family level. 

```{r composition-amplicon-example3, fig.width=12, fig.height=5, out.width="400px", fig.show="hold", warning=FALSE, message=FALSE, eval=TRUE}
library(phyloseq)
# merge at family level.

pseq.fam <- aggregate_top_taxa(ps1, top=10,"Family")

p.fam <- plot_composition(pseq.fam, sample.sort = NULL, 
                          otu.sort = NULL,
                          x.label = "empo_3", # sample type
                          plot.type = "barplot", 
                          verbose = FALSE) + 
  theme_bw() + scale_fill_brewer("Family", palette = "Paired")
# we can rotate x axis labels 
print(p.fam + theme(axis.text.x = element_text(angle = 90)))

```

## Plot relative abundance

```{r composition-amplicon-example4, fig.width=12, fig.height=5, out.width="400px", fig.show="hold", warning=FALSE, message=FALSE, eval=TRUE}

pseq.famrel <- microbiome::transform(pseq.fam, "compositional")

p.famrel <- plot_composition(pseq.famrel, sample.sort = NULL, otu.sort = NULL,
  x.label = "empo_3", plot.type = "barplot", verbose = FALSE)

print(p.famrel)

# further improvements can be done as follows  

p.famrel <- plot_composition(pseq.famrel, 
                             sample.sort = NULL, 
                             otu.sort = NULL, 
                             x.label = "empo_3", 
                             plot.type = "barplot", 
                             verbose = FALSE) + 
  guides(fill = guide_legend(ncol = 1)) + 
  labs(x = "Samples", 
       y = "Relative abundance",
       title = "Relative abundance data", 
       subtitle = "Subtitle",
       caption = "Caption text.") + 
     scale_fill_brewer("Family", palette = "Paired")

print(p.famrel)
```

Average by group

```{r composition-amplicon-example5, fig.width=6, fig.height=5, out.width="400px", fig.show="hold", warning=FALSE, message=FALSE, eval=TRUE}
# Averaged by group
p <- plot_composition(pseq.famrel,
                      average_by = "empo_3") + 
  guides(fill = guide_legend(ncol = 1)) + 
  labs(x = "Samples", 
       y = "Relative abundance",
       title = "Relative abundance data", 
       subtitle = "Subtitle",
       caption = "Caption text.") 
print(p + scale_fill_brewer("Family", palette = "Paired") + theme_bw())

```

# Heatmap composition  

```{r composition-amplicon-example6, fig.width=12, fig.height=5, out.width="400px", fig.show="hold", warning=FALSE, message=FALSE, eval=TRUE}

pseq.famlog <- microbiome::transform(pseq.fam, "log10")

p.famrel.heatmap <- plot_composition(pseq.famlog, 
                             sample.sort = NULL, 
                             otu.sort = NULL, 
                             x.label = "empo_3", 
                             plot.type = "heatmap", 
                             verbose = FALSE)

print(p.famrel.heatmap)
```

## Plot core taxa time trajectory
```{r}
library(dplyr)
# select core
ps <- moving_pictures
table(meta(ps)$sample_type, meta(ps)$host_subject_id)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
# Filter the data to include only gut samples from M3 subject
ps.m3 <- subset_samples(ps, sample_type == "stool" & host_subject_id == "M3") 
#print(ps.m3)

ps.m3.rel <- microbiome::transform(ps.m3, "compositional")
pseq.core <- core(ps.m3.rel, detection = 0.001, prevalence = .95)

ps.stool.df <- psmelt(pseq.core)
#head(ps.stool.df)

# add genus name to ASVid
ps.stool.df <- ps.stool.df %>% 
  mutate(asv_gen= paste0(OTU, "-",Genus))

ps.stool.rel.plot <- ggplot(ps.stool.df) + 
  geom_line(aes(days_since_experiment_start, 
                Abundance, color = asv_gen)) +
  theme_bw() + 
  theme(legend.position="top") + 
  xlab("Days since experiment start") + 
  scale_color_brewer(palette = "Paired") +
  guides(col = guide_legend(ncol = 6, nrow = 6))

ps.stool.rel.plot
```

