---
title: "Shared core with venn diagram"
author: "Sudarshan Shetty, Leo Lahti, et al."
bibliography: 
- bibliography.bib
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
    toc_float: true
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    use_bookdown: false
    highlight: haddock
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - atlas}
  %\usepackage[utf8]{inputenc}
-->


In may instances one would wish to compare core taxa that are shared between multiple groups. Here, we demonstrate how this can be achieved by `microbiome` and [eulerr](https://cran.r-project.org/web/packages/eulerr/vignettes/introduction.html). The test data is stored in the [microbiomeutilities](https://microsud.github.io/microbiomeutilities/articles/microbiomeutilities.html) R package and the original source of data is [Zackular et al., 2014: The Gut Microbiome Modulates Colon Tumorigenesis](http://mbio.asm.org/content/4/6/e00692-13).   

Load packages. 
```{r pkg}
#install.packages("eulerr") # If not installed
library(eulerr)
library(microbiome)
#devtools::install_github('microsud/microbiomeutilities')
library(microbiomeutilities)
```


Get data  
```{r data1, warning=FALSE, message=FALSE}
data("zackular2014")
pseq <- zackular2014
```


We will aim to identify shared core taxa between nationalities.  
```{r check-var}
# simple way to count number of samples in each group
table(meta(pseq)$DiseaseState, useNA = "always")
```
There are 30 CRC and Healthy samples while 28 CRC samples.  

```{r fix-pseq}

# convert to relative abundances
pseq.rel <- microbiome::transform(pseq, "compositional")

```

Make a list of DiseaseStates.    

```{r get-grps}

disease_states <- unique(as.character(meta(pseq.rel)$DiseaseState))
print(disease_states)

```

Write a `for loop` to go through each of the `disease_states` one by one and combine identified core taxa into a list.  
```{r get-core}

list_core <- c() # an empty object to store information

for (n in disease_states){ # for each variable n in DiseaseState
    #print(paste0("Identifying Core Taxa for ", n))
    
    ps.sub <- subset_samples(pseq.rel, DiseaseState == n) # Choose sample from DiseaseState by n
    
    core_m <- core_members(ps.sub, # ps.sub is phyloseq selected with only samples from g 
                           detection = 0.001, # 0.001 in atleast 90% samples 
                           prevalence = 0.75)
    print(paste0("No. of core taxa in ", n, " : ", length(core_m))) # print core taxa identified in each DiseaseState.
    list_core[[n]] <- core_m # add to a list core taxa for each group.
    #print(list_core)
}

```

```{r printlist, eval=FALSE}

print(list_core)

```


```{r plot-venn, fig.width=6, fig.height=6, fig.align='center'}
# Specify colors and plot venn
# supplying colors in the order they appear in list_core
mycols <- c(nonCRC="#d6e2e9", CRC="#cbf3f0", H="#fcf5c7") 
plot(venn(list_core),
     fills = mycols)

```

```{r printlist-2, eval=TRUE}

print(list_core)

```
Here, we see that the ids are only names without taxonomic information. This is because the rownames of `otu_table` and `tax_table` are usually IDs or sequences.  
The `format_to_besthit()` function in `microbiomeutilites` can be useful which add best taxonomic information that is available as demonstrated below. 


```{r}

# use the pseq.rel object created at the begening of this tutorial. 
taxa_names(pseq.rel)[1:5]

```
The taxa names are ids.  

```{r}

# first remove "d__"
taxa_names(pseq.rel) <- gsub( "d__","", taxa_names(pseq.rel))
taxa_names(pseq.rel)[1:5]
# format names
pseq.rel.f <- format_to_besthit(pseq.rel)
# check names
taxa_names(pseq.rel.f)[1:5]
```

Run the `for loop` to go through each of the `disease_states` one by one and combine identified core taxa into a list.  

```{r get-core-2}

list_core <- c() # an empty object to store information

for (n in disease_states){ # for each variable n in DiseaseState
    #print(paste0("Identifying Core Taxa for ", n))
    
    ps.sub <- subset_samples(pseq.rel.f, DiseaseState == n) # Choose sample from DiseaseState by n
    
    core_m <- core_members(ps.sub, # ps.sub is phyloseq selected with only samples from g 
                           detection = 0.001, # 0.001 in atleast 90% samples 
                           prevalence = 0.75)
    print(paste0("No. of core taxa in ", n, " : ", length(core_m))) # print core taxa identified in each DiseaseState.
    list_core[[n]] <- core_m # add to a list core taxa for each group.
    #print(list_core)
}

```
This is more useful information.  
```{r}
print(list_core)
```
Note: There are limitations to the number of groups that can be plotted with the `eulerr::venn()`. Check [eulerr](https://cran.r-project.org/web/packages/eulerr/vignettes/introduction.html) documentation for more detailed information.  



