---
title: "post_hoc"
author: "Leo Lahti, Sudarshan Shetty et al."
bibliography: 
- bibliography.bib
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
    toc_float: true
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    use_bookdown: false
    highlight: haddock
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - comparisons}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


## Multiple pairwise-comparison between groups 



Load example data:

```{r post_hoc_libraries, warning=FALSE, message=FALSE}
# Load libraries
library(microbiome)
library(ggplot2)
library(dplyr)

data("dietswap")
pseq <- dietswap # Rename the example data
```


Visually compare Akkermansia abundance between "DI", "ED" and "HE"

```{r post_hoc_boxplot_kruskal, warning=FALSE, message=FALSE}
p <- boxplot_abundance(pseq, x = "group", y = "Akkermansia") + scale_y_log10()
print(p)
```


### Test statistical significance by Kruskal-Wallis


```{r kruskal, message=FALSE, error=FALSE, warning=FALSE}

library(tidyr)

# Analyse specific taxa
tax <- "Akkermansia"

# Pick the signal (abundance) for this tax
sample_data(pseq)$signal <- get_sample(pseq, tax)


attach(meta(pseq))
res<- pairwise.wilcox.test(signal, group, p.adjust.method = 'BH')
resK<- kruskal.test(signal, group) # source: http://www.sthda.com/english/wiki/kruskal-wallis-test-in-r#what-is-kruskal-wallis-test
# Show the results
print(resK)
```


From the output of the Kruskal-Wallis test, if there is a significant difference between groups, but we don’t know which pairs of groups are different, then we shall move to:


#### Test statistical significance with common post hoc as follows:

##### 1. Pairwise Wilcoxon tests with multiple testing correction, p.adjust.method is Benjamini-Hochberg

```{r post_hoc_Pairwise_Wilcoxon_tests, message=FALSE, error=FALSE, warning=FALSE}

library(tidyr)

# Analyse specific taxa
tax <- "Akkermansia"

# Pick the signal (abundance) for this tax
sample_data(pseq)$signal <- get_sample(pseq, tax)


attach(meta(pseq))
res<- pairwise.wilcox.test(signal, group, p.adjust.method = 'BH')

# Show the results
print(res)
```






#### 2. Dunn's test, p.adjust.method is Benjamini-Hochberg

```{r post_hoc_dunn.test, message=FALSE, error=FALSE, warning=FALSE}

library(dunn.test)

# Analyse specific taxa
tax <- "Akkermansia"

# Pick the signal (abundance) for this tax
sample_data(pseq)$signal <- get_sample(pseq, tax)


attach(meta(pseq))
res<- dunn.test::dunn.test(signal, group, method = 'bh') # source:https://cran.r-project.org/web/packages/dunn.test/dunn.test.pdf

print(res)

```




#### 3. Pairwise Multiple Comparison of Mean Ranks (PMCMR)

```{r post_hoc_PMCMR, message=FALSE, error=FALSE, warning=FALSE}

library(PMCMR)

# Analyse specific taxa
tax <- "Akkermansia"

# Pick the signal (abundance) for this tax
sample_data(pseq)$signal <- get_sample(pseq, tax)


attach(meta(pseq))
res<- PMCMR::posthoc.kruskal.nemenyi.test(signal, group,dist='Chisq',p.adjust.methods=  'BH') # source:https://cran.r-project.org/web/packages/PMCMR/vignettes/PMCMR.pdf

# Show the results
print(res)

```


#### 4. Conover-Iman Test of Multiple Comparisons Using Rank Sums, p.adjust.method is Benjamini-Hochberg

```{r post_hoc_Conover-Iman, message=FALSE, error=FALSE, warning=FALSE}

library(conover.test)

# Analyse specific taxa
tax <- "Akkermansia"

# Pick the signal (abundance) for this tax
sample_data(pseq)$signal <- get_sample(pseq, tax)

# source:https://cran.r-project.org/web/packages/conover.test/conover.test.pdf

attach(meta(pseq))
res<- conover.test::conover.test(signal, group, method = 'bh')
  



```
