---
title: "Group comparisons at the level of individual taxonomic groups"
author: "Author"
date: "`r date()`"
output: md_document
---

```{r init, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
theme_set(theme_bw(20))
opts_chunk$set(fig.width=10, fig.height=10, message=FALSE, warning=FALSE)
opts_chunk$set(dev="CairoPNG")
padj.th <- 0.25
```

# ANCOM analysis

Significant taxa (regarding the grouping). Only the hits with adjusted p<`r padj.th` are shown. Adjustment is done per group based on ANCOM defaults.

The log10 fold changes have been calculated from CLR-transformed abundances.

```{r deseq2, echo=FALSE, warning=FALSE, message=FALSE}
tab <- subset(res.ancom, pval.ancom < 0.05) %>% arrange(desc(abs(log10FC)))
tab$relab.cond <- round(tab$relab.cond, 2)
tab$relab.control <- round(tab$relab.control, 2)
kable(tab)
```


Illustration of the significant genera. CLR transformed abundances (just first 3 shown).

```{r tops, echo=FALSE, warning=FALSE, message=FALSE, fig.show="hold", out.width = "20%", fig.height=6, fig.width=5}
library(ggbeeswarm)
theme_set(theme_bw(20))
df <- meta(pseq)
ps <- transform(pseq, "clr")
tab <- subset(res.ancom, pval.ancom < 0.05) %>% arrange(desc(abs(log10FC)))
for (i in 1:min(nrow(tab), 3)) {
  tax <- as.character(res.ancom[i, "OTU"])
  group <- as.character(res.ancom[i, "Grouping"]    )
  df <- meta(ps)
  df$Group <- as.factor(df[, gsub("/", "\\.", group)])
  df$Abundance <- abundances(ps)[tax, ]
  df <- subset(df, !is.na(Group))
  p <- ggplot(df, aes(x = Group, y = Abundance)) +
         #geom_boxplot() +
	 geom_beeswarm() +
	 #geom_jitter(width = 0.25, alpha = 0.5) + 	 
	 # scale_y_log10() +
	 #coord_flip() +
	 theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) + 
	 labs(x = "",
	      y = "Abundance (CLR)",
	      title = paste(tax),
	      subtitle = group)
  print(p)
}
```



